Case Study: 16S microbiome data (Schubert et al., 2014)

### 2018-07-06 datasets/microbiome/microbiome-enigma.Rmd
Add uninformative covariate analyses to RNAseq

# From their paper, we expect SO4 to have little correlation, pH
# to have high correlations, and Al to be in the middle
so4_result_file <- file.path(resdir, "enigma-so4-otus-results.rds")
so4_bench_file <- file.path(sbdir, "enigma-so4-otus-benchmark.rds")
so4_bench_file_abun <- file.path(sbdir, "enigma-so4-otus-abun-benchmark.rds")

ph_result_file <- file.path(resdir, "enigma-ph-otus-results.rds")
ph_bench_file <- file.path(sbdir, "enigma-ph-otus-benchmark.rds")
ph_bench_file_abun <- file.path(sbdir, "enigma-ph-otus-abun-benchmark.rds")

al_result_file <- file.path(resdir, "enigma-al-otus-results.rds")
al_bench_file <- file.path(sbdir, "enigma-al-otus-benchmark.rds")
al_bench_file_abun <- file.path(sbdir, "enigma-al-otus-abun-benchmark.rds")
```

@@

@@

### Multiple-Testing Correction

@@

### Benchmark Metrics

@@

```{r, ph-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(ph_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = ph_bench_file_abun)
} else {
    sb <- readRDS(ph_bench_file_abun)
}
```
### Benchmark Metrics

@@

@@

@@

### Multiple-Testing Correction

@@

```{r, so4-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(so4_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = so4_bench_file_abun)
} else {
    sb <- readRDS(so4_bench_file_abun)
}
```
### Benchmark Metrics

@@

@@

@@

### Multiple-Testing Correction

@@

```{r, al-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(al_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = al_bench_file_abun)
} else {
    sb <- readRDS(al_bench_file_abun)
}
```
### Benchmark Metrics

@@

# Response comparison

@@

```{r}
plotMethodRanks(c(so4_bench_file, so4_bench_file_abun,
                  ph_bench_file, ph_bench_file_abun,
                  al_bench_file, al_bench_file_abun), 
                colLabels = c("SO4", "SO4-abun", "pH", "ph-abun",
                              "Al", "Al-abun"), 
                alpha = 0.10, xlab = "Comparison")
```

>>>>

# From their paper, we expect SO4 to have little correlation, pH
# to have high correlations, and Al to be in the middle
so4_result_file <- file.path(resdir, "enigma-so4-otus-results.rds")
so4_bench_file <- file.path(sbdir, "enigma-so4-otus-benchmark.rds")
so4_bench_file_abun <- file.path(sbdir, "enigma-so4-otus-abun-benchmark.rds")
so4_bench_file_uninf <- file.path(sbdir, "enigma-so4-otus-uninf-benchmark.rds")

ph_result_file <- file.path(resdir, "enigma-ph-otus-results.rds")
ph_bench_file <- file.path(sbdir, "enigma-ph-otus-benchmark.rds")
ph_bench_file_abun <- file.path(sbdir, "enigma-ph-otus-abun-benchmark.rds")
ph_bench_file_uninf <- file.path(sbdir, "enigma-ph-otus-uninf-benchmark.rds")

al_result_file <- file.path(resdir, "enigma-al-otus-results.rds")
al_bench_file <- file.path(sbdir, "enigma-al-otus-benchmark.rds")
al_bench_file_abun <- file.path(sbdir, "enigma-al-otus-abun-benchmark.rds")
al_bench_file_uninf <- file.path(sbdir, "enigma-al-otus-uninf-benchmark.rds")
```

@@

Add random (uninformative) covariate.

```{r}
set.seed(44318)
res$rand_covar <- rnorm(nrow(res))
```

@@

#### Random

```{r, ph-rand-hist, fig.width=10, fig.height=3.2}
strat_hist(res, pvalue="pval", covariate="rand_covar", maxy=8, binwidth=0.05)
```

```{r, ph-rand-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res, pvalue="pval", covariate="rand_covar")
```

### Multiple-Testing Correction - ubiquity

@@

### Benchmark Metrics - ubiquity

@@

```{r, ph-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(ph_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = ph_bench_file_abun)
} else {
    sb <- readRDS(ph_bench_file_abun)
}
```
### Benchmark Metrics - mean abundance

@@

### Multiple-Testing Correction - random

Let's use `rand_covar` as our `ind_covariate`.

```{r, ph-labelcovariate-rand}
res <- dplyr::mutate(res, ind_covariate = rand_covar)
```

```{r, ph-sb-rand, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(ph_bench_file_uninf)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = ph_bench_file_uninf)
} else {
    sb <- readRDS(ph_bench_file_uninf)
}
```
### Benchmark Metrics - random

```{r, ph-metrics-rand, eval=TRUE}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, ph-plot-rand, width=15, height=15, eval=TRUE}
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
```
```{r}
covariateLinePlot(sb, alpha = 0.05, covname = "ind_covariate")
```

@@

Add random (uninformative) covariate.

```{r}
set.seed(27349)
res$rand_covar <- rnorm(nrow(res))
```

@@

#### Random

```{r, so4-rand-hist, fig.width=10, fig.height=3.2}
strat_hist(res, pvalue="pval", covariate="rand_covar", maxy=8, binwidth=0.05)
```

```{r, so4-rand-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res, pvalue="pval", covariate="rand_covar")
```

### Multiple-Testing Correction - ubiquity

@@

```{r, so4-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(so4_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = so4_bench_file_abun)
} else {
    sb <- readRDS(so4_bench_file_abun)
}
```
### Benchmark Metrics - mean abundance

@@

### Multiple-Testing Correction - random

Let's use `rand_covar` as our `ind_covariate`.

```{r, so4-labelcovariate-rand}
res <- dplyr::mutate(res, ind_covariate = mean_abun_present)
```

```{r, so4-sb-rand, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(so4_bench_file_uninf)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = so4_bench_file_uninf)
} else {
    sb <- readRDS(so4_bench_file_uninf)
}
```
### Benchmark Metrics - random

```{r, so4-metrics-rand, eval=TRUE}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, so4-plot-rand, width=15, height=15, eval=TRUE}
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
```
```{r}
covariateLinePlot(sb, alpha = 0.05, covname = "ind_covariate")
```

@@

Add random (uninformative) covariate.

```{r}
set.seed(19474)
res$rand_covar <- rnorm(nrow(res))
```

@@

#### Random

```{r, al-rand-hist, fig.width=10, fig.height=3.2}
strat_hist(res, pvalue="pval", covariate="rand_covar", maxy=8, binwidth=0.05)
```

```{r, al-rand-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res, pvalue="pval", covariate="rand_covar")
```

### Multiple-Testing Correction - ubiquity

@@

```{r, al-sb-abun, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(al_bench_file_abun)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = al_bench_file_abun)
} else {
    sb <- readRDS(al_bench_file_abun)
}
```
### Benchmark Metrics - mean abundance

@@

### Multiple-Testing Correction - random

Let's use `rand_covar` as our `ind_covariate`.

```{r, al-labelcovariate-rand}
res <- dplyr::mutate(res, ind_covariate = rand_covar)
```

```{r, al-sb-rand, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(al_bench_file_uninf)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    saveRDS(sb, file = al_bench_file_uninf)
} else {
    sb <- readRDS(al_bench_file_uninf)
}
```
### Benchmark Metrics - random

```{r, al-metrics-rand, eval=TRUE}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, al-plot-rand, width=15, height=15, eval=TRUE}
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
```
```{r}
covariateLinePlot(sb, alpha = 0.05, covname = "ind_covariate")
```

# Response/Covariate comparison

@@

```{r, width = 15, height = 10}
plotMethodRanks(c(so4_bench_file, so4_bench_file_abun, so4_bench_file_uninf,
                  ph_bench_file, ph_bench_file_abun, ph_bench_file_uninf,
                  al_bench_file, al_bench_file_abun, al_bench_file_uninf), 
                colLabels = c("SO4-ubiquity", "SO4-abun", "SO4-uninf",
                              "pH-ubiquity", "ph-abun", "pH-uninf",
                              "Al-ubiquity", "Al-abun", "Al-uninf"),
                alpha = 0.10, xlab = "Comparison")
```

### 2018-07-06 datasets/microbiome/microbiome-papa.Rmd
Add uninformative covariate analyses to RNAseq

otu_result_file <- file.path(resdir, "papa-otus-results.rds")
otu_bench_file <- file.path(sbdir, "papa-otus-benchmark.rds")
gns_result_file <- file.path(resdir, "papa-genus-results.rds")
gns_bench_file <- file.path(sbdir, "papa-genus-benchmark.rds")
gns_bench_file_log_ubi <- file.path(resdir, "papa-genus-log-ubiquity-benchmark.rds")
gns_bench_file_abun <- file.path(resdir, "papa-genus-abun-benchmark.rds")
```

@@

@@

@@

@@

@@

```{r}
plotMethodRanks(c(otu_bench_file, gns_bench_file, 
                  gns_bench_file_log_ubi,gns_bench_file_abun), 
                colLabels = c("OTU", "Genus", "Genus-log", "Genus-abun"), 
                alpha = 0.10, xlab = "Comparison")
```

>>>>

otu_result_file <- file.path(resdir, "papa-otus-results.rds")
otu_bench_file <- file.path(sbdir, "papa-otus-benchmark.rds")
gns_result_file <- file.path(resdir, "papa-genus-results.rds")
gns_bench_file <- file.path(sbdir, "papa-genus-benchmark.rds")
gns_bench_file_log_ubi <- file.path(sbdir, "papa-genus-log-ubiquity-benchmark.rds")
gns_bench_file_abun <- file.path(sbdir, "papa-genus-abun-benchmark.rds")
gns_bench_file_uninf <- file.path(sbdir, "papa-genus-uninf-benchmark.rds")
```

@@

Add random (uninformative) covariate to test at the genus level.

```{r}
set.seed(33128)
res$rand_covar <- rnorm(nrow(res))
```

@@

### Random

```{r, papa-genus-rand, fig.width=10, fig.height=3.2}
strat_hist(res, pvalue="pval", covariate="rand_covar", maxy=8, binwidth=0.05)
```

```{r, papa-genus-rand-scatter, fig.width=4.5, fig.height=3.5}
rank_scatter(res, pvalue="pval", covariate="rand_covar")
```

## Multiple-Testing Correction - ubiquity

@@

## Benchmark Metrics - ubiquity

@@

## Multiple-Testing Correction - random

Let's use the random (uninformative) covariate as our `ind_covariate`.

```{r, papa-genus-rand-labelcovariate}
res <- dplyr::mutate(res, ind_covariate = rand_covar)
```

Then, we'll construct the `SummarizedBenchmark` object, which will run
the functions specified in each method (these are actually sourced in from the
helper scripts). 
```{r, papa-genus-sb-rand, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(gns_bench_file_uninf)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb_rand <- buildBench(bd, data = res, ftCols = "ind_covariate")
    metadata(sb_rand)$data_download_link <-
                   "https://zenodo.org/record/840333/files/ibd_alm_results.tar.gz"
    saveRDS(sb_rand, file = gns_bench_file_uninf)
} else {
    sb_rand <- readRDS(gns_bench_file_uninf)
}
```

## Benchmark Metrics - random

```{r, papa-genus-rand-metrics, eval=TRUE}
assayNames(sb_rand) <- "qvalue"
sb_rand <- addDefaultMetrics(sb_rand)
```

```{r, papa-genus-rand-plot, width=15, height=15, eval=TRUE}
rejections_scatter(sb_rand, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb_rand, covariate="ind_covariate", supplementary=FALSE)
```

```{r, papa-genus-rand-overlap}
plotFDRMethodsOverlap(sb_rand, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```
```{r}
methods <- c( "lfdr", "ihw-a10", "bl-df03", "qvalue", "bh", "bonf" )
plotCovariateBoxplots(sb_rand, alpha=0.1, nsets=6, methods=methods)
```
```{r}
covariateLinePlot(sb_rand, alpha = 0.05, covname = "ind_covariate")
```

@@

```{r}
plotMethodRanks(c(otu_bench_file, gns_bench_file, 
                  gns_bench_file_log_ubi,gns_bench_file_abun,
                  gns_bench_file_uninf), 
                colLabels = c("OTU-ubiquity", "Genus-ubiquity", "Genus-log", "Genus-abun", "Genus-uninf"), 
                alpha = 0.10, xlab = "Comparison")
```

### 2018-07-06 datasets/microbiome/microbiome-schubert.Rmd
Add uninformative covariate analyses to RNAseq

otu_result_file <- file.path(resdir, "schubert-otus-results.rds")
otu_bench_file <- file.path(sbdir, "schubert-otus-benchmark.rds")
otu_bench_file_abun <- file.path(sbdir, "schubert-otus-abun-benchmark.rds")
```

@@

@@

```{r}
plotMethodRanks(c(otu_bench_file, otu_bench_file_abun), 
                colLabels = c("OTU","OTU-abun"), 
                alpha = 0.10, xlab = "Comparison")
```

>>>>

otu_result_file <- file.path(resdir, "schubert-otus-results.rds")
otu_bench_file <- file.path(sbdir, "schubert-otus-benchmark.rds")
otu_bench_file_abun <- file.path(sbdir, "schubert-otus-abun-benchmark.rds")
otu_bench_file_uninf <- file.path(sbdir, "schubert-otus-uninf-benchmark.rds")
```

@@

## Multiple-Testing Correction - random

Let's use `rand_covar` as our `ind_covariate`.

```{r, schubert-labelcovariate-rand, eval=TRUE}
res <- dplyr::mutate(res, ind_covariate = rand_covar)
```

```{r, schubert-sb-rand, results="hide", message=FALSE, eval=TRUE}
if (!file.exists(otu_bench_file_uninf)) {
    bd <- initializeBenchDesign()
    bd <- dropBMethod(bd, "ashq")
    sb <- buildBench(bd, data = res, ftCols = "ind_covariate")
    metadata(sb)$data_download_link <-
                   "https://zenodo.org/record/840333/files/cdi_schubert_results.tar.gz"
    saveRDS(sb, file = otu_bench_file_uninf)
} else {
    sb <- readRDS(otu_bench_file_uninf)
}
```

## Benchmark Metrics - random

```{r, schubert-metrics-rand, eval=TRUE}
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

```{r, schubert-plot-rand, width=15, height=15, eval=TRUE}
rejections_scatter(sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```
```{r}
covariateLinePlot(sb, alpha = 0.05, covname = "ind_covariate")
```

@@

```{r}
plotMethodRanks(c(otu_bench_file, otu_bench_file_abun, otu_bench_file_uninf), 
                colLabels = c("OTU-ubiquity","OTU-abun", "OTU-uninf"), 
                alpha = 0.10, xlab = "Comparison")
```
